---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
terra::terraOptions(progress = 0)
```

# bsspread: Biosecurity Population Spread Modelling

<!-- badges: start -->
[![Last commit](https://img.shields.io/github/last-commit/cebra-analytics/bsspread.svg)](https://github.com/cebra-analytics/bsspread/commits/main)
<!-- badges: end -->

The *bsspread* package provides a collection of workflow components implemented in *R* (<https://www.r-project.org/>) as *S3* object classes, which encapsulate functionality for building and running stochastic simulations that model the population dynamics and spatio-temporal spread of biosecurity threats. Population spread simulation models are constructed by building and linking workflow components as follows:

1. Define the spatial representation and region of interest for the study.
1. Configure the population representation and dynamics for the biosecurity threat.
1. Specify how each simulation is initialised.
1. Assemble one or more dispersal models for each mode or vector of threat spread.
1. Configure and run one or more stochastic spatio-temporal simulations and save their results.

## Workflow components

This section further describes the workflow components for building population spread simulation models.

### Region

The region of interest for the sutdy may be configured via the *Region* object class. The spatial representation may be configured as either:

1. A grid-based raster layer (GeoTIFF file), whereby model processes are configured and simulated using the active (non-NA) cell locations.
1. A network of locations or patches defined via a table of longitude and latitude coordinates (CSV file).
1. A spatially implicit area.

When defined by a grid-based raster layer, the region can be configured for two-tier dispersal, whereby long-distance dispersal is calculated using a courser (aggregate) spatial resolution for destinations outside of an inner radius from each threat-occupied location, within which local dispersal is calculated using the original spatial resolution of the region.

The *Region* class provides functions for calculating distances and directions between (local and aggregate) cells or patches, which is utilised by dispersal model components (see below).

The *Region* class also provides functionality for configuring and calculating modified dispersal distances for grid-based permeability layers or network-based weighted paths (graphs), which is also optionally utilised by dispersal model components (see below).

### Population models

The population representation and dynamics of the biosecurity threat may be configured via one of the following object classes, which are based on (or *inherit*) the *Population* class:

1. *PresencePopulation*: Models presence-only populations.
1. *UnstructPopulation*: Models unstructured populations and their growth.
1. *StagedPopulation*: Models stage or age-based populations and their reproduction and survival dynamics.

Presence-only populations are modelled via a binary indication of threat presence or absence at each location at each simulation time step. An optional *spread delay* parameter may be configured to account for the time it takes for a threat species to have the ability to spread after it has established in each new location (e.g. seeding plants). This representation provides a simple approach that may be suitable for modelling spread for some threats, and are particularly useful when the parameters required for growth or population dynamics are unknown (Bradhurst et al., 2021; Jongejans, Skarpaas, & Shea, 2008).

Unstructured populations are modelled via a numeric population size at each location at each simulation time step (Jongejans, Skarpaas, & Shea, 2008). A *growth* parameter configures the population growth rate or lambda (e.g. 1.2 for 20% growth per time step). Capacity-limited growth (Ricker, 1958) may be modelled by specifying the carrying *capacity* of the threat at each location for grid & network-based models, or for each unit area occupied in spatially implicit models.

Stage or age-based populations are modelled via a numeric population sizes for each life-stage or age of the threat at each location at each simulation time step. The reproduction and survival dynamics of the threat species are modelled via stage or age transition rates specified via a *growth* matrix (Lefkovitch, 1965; Leslie, 1945). Again, the capacity-limited growth (transition rates) may be configured by specifying the carrying *capacity*, as well as the life stages or ages that are applicable for capacity-limited dynamics (García Adeva, Botha, & Reynolds, 2012; Jongejans, Skarpaas, & Shea, 2008).

Either population model (presence-only, unstructured, or staged) may be optionally configured with the likelihood or probability that the pest will establish on arrival for each grid or network-based dispersal destination (Bradhurst et al., 2021; García Adeva, Botha, & Reynolds, 2012; Jongejans, Skarpaas, & Shea, 2008). This establishment probability may also be configured to vary temporally within the simulated model. Likewise, the growth or transition rates, as well as the carrying capacity (in unstructured or staged populations) may be configured to vary spatially, temporally, or both (spatio-temporally) within the simulated model,

### Initialisation

The generation of the initial population for each model simulation may be configured via the  *Initializer* class using either of the following approaches (Bradhurst et al., 2021; García Adeva, Botha, & Reynolds, 2012):

1. Pre-defined initial population values at specified locations.
1. Stochastically generate an initial population for each simulation (replicate) at a location chosen via sampling using incursion weightings specified for each location.
1. Stochastically generate an initial population at one or more (or no) locations selected via sampling using arrival probabilities specified at each location. The stochastically generated threat arrivals may optionally continue to be generated at specified simulation time intervals.

Both stochastic generation options also utilise functionality provided by the *Incursions* class. Initial population sizes for unstructured and staged models are generated via Poison sampling using specified means. Initial distributions of staged populations are determined stochastically via expected stage/age ratios, given the stages/ages specified for initialisation.

### Dispersal models

One or more dispersal models may be assembled and configured for each mode or vector of threat spread. The various ways organisms may spread, described in Gippet et al. (2019) and Wilson et al. (2009), include:

- Leading edge spread (or 'diffusion') from established occupancy to nearby locations.
- Corridor spread along geographical features such as rivers, roads, etc.
- Natural long-range spread (commonly referred to as 'jumps') by wind, water currents, migratory animals, etc.
- Human-mediated spread (also referred to as 'jumps') by transport, trade, tourism, etc. 

Assembling and configuring an appropriate dispersal model for a particular mode of spread may be achieved via the following object classes:

1. *Dispersal*: A generic *base* class with dispersal functionality for spatially explicit (grid and network-based) spread of threat *populations* (Bradhurst et al., 2021; García Adeva, Botha, & Reynolds, 2012; Jongejans, Skarpaas, & Shea, 2008; Robinet et al. 2012), including configuration for:
    - The *proportion* of threat individuals dispersing at each time step.
    - The number of dispersal *events* from each occupied location at each time step.
    - *Density dependent* dispersal for spread that increases as the number of individuals approach *capacity* (e.g. in response to overcrowding).
    - The likelihood of dispersal given destination distances (via a *distance function* or *kernel*).
    - The likelihood of dispersal given destination directions (via a *direction function* or *kernel*).
    - The likelihood of dispersal given other destination attributes (e.g. food or habitat) or 'attractiveness' (via *attractors*).
    - The *permeability* of the (grid-based) landscape or connectivity of (network-based) locations.

2. *Kernels*: A class for generating distance or direction functions (or kernels) for calculating the (relative) probability of dispersal given destination distances or directions, used by *Dispersal* models and inherited class objects. The kernel functions include the following commonly used probability distributions (Carrasco et al., 2010-1; Jongejans, Skarpaas, & Shea, 2008; Shaw, 1995):
    - Uniform
    - Beta (with shift)
    - Cauchy
    - Exponential (negative exponential)
    - Gaussian (normal)
    - Lognormal
    - Weibull
    
    Also included is a function for defining the probability of distances or directions from a table of values.

3. *DirectionKernel*: A class extending the functionality of the *Kernels* class specifically for generating direction functions (or kernels), which enables:
    - Directions to be defined in degrees (0-360) using trigonometric angles or navigational directions.
    - The Beta distribution to be shifted or rotated (e.g. for dominant winds).

4. *Gravity*: A class *inherited* from the *Dispersal* class configured for network gravity models. Gravity models provide mechanisms for biased dispersal patterns based on a combination of distance between locations and 'attraction' between origins and destinations, such as population, environmental, and other conditions suitable for the threat organism or its hosts, or factors influenced by human behaviour. Gravity models are analogous to Newton’s gravitational law, whereby the attraction decreases with distance (Bossenbroek, Kraft, & Nekola, 2001; Carrasco et al., 2010-2; Crespo-Pérez et al., 2011; Muirhead et al., 2006). They are typically expressed as a function in the form:

    $dispersal_{ij}=\frac{f(attractor1_{ij},attractor2_{ij},...)}{distance_{ij}^\beta}$

    where a function $f$ of *attractor* variables $attractor1$, $attractor2$, ... (often a simple product), combined with the inverse distance to the power of a constant $\beta$ (often 1), is used to calculate the dispersal between origin and destination locations $i$ and $j$ respectively.

5. *Attractor*: A class for encapsulating attractor variables, which can be used in spatially explicit (grid and network-based) dispersal models. Examples of attractor variables include (Bossenbroek, Kraft, & Nekola, 2001; Carrasco et al., 2010-2; Crespo-Pérez et al., 2011; Muirhead et al., 2006; Schneider, Ellis, & Cummings, 1998):
    - Population size or density of threat organisms, host species, or human settlements.
    - Carrying capacity or suitability.
    - Environmental or climatic conditions.
    - Human transport movement numbers or capacity between origins and destinations.
    - Availability/quantity of host material, such as wood.

6. *Permeability*: a class to represent the permeability of grid-based landscapes, or location (node/patch) connectivity in network models. Both approaches are implemented via *igraph* (Antonov et al., 2023; Csárdi & Nepusz, 2006; Csárdi et al., 2026):
    - Grid-based permeability uses approaches derived from Etherington (2016), whereby the permeability layer values (0-1) of cells is used to scale the actual distance between adjacent cells when applying a dispersal *distance function* (as above). For example, a permeability value of 0.5 results in an effective distance twice that of the actual distance, a value of 0 prevents spread to or through a cell, whereas a value of 1 does not modify the effective distance.
    - Network-based connectivity is defined by listing pairs of connected nodes and the weight (0-1) of each connection, thereby limiting the interconnectivity of the network to the pairs listed, as well as modifying the effective distance between connected nodes (as above).

7. *Diffusion*: A class *inherited* from the *Dispersal* class configured for diffusive spread, which is implemented differently dependent on the spatial (*region*) and *population* configuration:
    - For spatially explicit (grid and network-based) models, diffusion simulates the local spread of populations into neighbouring locations at a speed specified by a *diffusion rate* (Bradhurst et al., 2021; Jongejans, Skarpaas, & Shea, 2008). The optional configuration of *direction function*, *attractors*, and/or *permeability* will thus potentially result in asymmetric diffusion.
    - For spatially implicit models, diffusion is also defined by a speed specified by a *diffusion rate*. For presence-only populations a simple radial diffusion is applied, whereby the implicit area occupied expands via a circle with a linear increasing radius (up to a maximum area when specified by the *region*). For unstructured and stage-based population models, a *reaction-diffusion* model is applied, whereby capacity-limited growth interplays with radial expansion, which slows down if the population density approaches capacity (Andow et al., 1990; Fisher, 1937; Okubo & Kareiva, 2001; Robinet et al. 2012; Skellam, 1951).

8. *AreaSpread*: A class *inherited* from the *Dispersal* class configured for spatially implicit area spread, whereby the simulated occupied area grows in proportion to the size of an unstructured or stage-based population given its capacity per unit area. When a maximum total area occupied is specified (via the *region*), then both the population and its area of occupancy will follow a typical capacity-limited sigmoid growth curve (Andow et al., 1990; Bradhurst et al., 2021; Jongejans, Skarpaas, & Shea, 2008; Robinet et al. 2012).

### Simulator

A *Simulator* class object is utilised to run one or more stochastic simulations of the population spread model utilising the (above) model components (Bradhurst et al., 2021; García Adeva, Botha, & Reynolds, 2012). The simulator is configured with:

- A *Region* class object to define the area of interest.
- Configuration for the simulation time steps, including:
    - The number of time steps to be simulated per model simulation.
    - The duration that each time step represents (e.g. 3 months).
    - The interval in time steps for collating results for all region locations.
- The number of model simulations replicates (albeit with stochastic variation) to run. 
- The *Initializer* object configured to initialise the threat population for each simulation.
- The *Population* or inherited class object configured with growth dynamics, carrying capacities, and establishment probabilities (when applicable).
- One or more *Dispersal* or inherited class objects configured for each mode or vector of threat spread, including distance and/or direction functions (kernels), attractors, and/or landscape permeability or network connectivity (when applicable).
- A *user function* for implementing custom functionality applied to the simulated population at each time step.

The simulator creates a *Results* class object to collate, generate, and save simulation results. Summary statistics (mean and standard deviations across replicates) are collated when more than one replicate is run. At each simulation time step the model processes are run in the following order:

1. Apply population growth or transitions.
1. Apply spread for each dispersal model (in order).
1. Apply the user function (when configured).
1. Collate the simulation results.

When all simulations are complete the simulator returns the *Result* class object with the collated simulation results.

## Installation

You can install the latest version of *bsspread* from [GitHub](https://github.com/) with:

``` r
remotes::install_github("cebra-analytics/bsspread")
```

## Example

The following example assembles and runs a population spread simulation model for a Hawkweed species (*Hieracium pilosella*), an exotic weed for Australia, which approximately reproduces the spread model described in Williams, Hahs, & Morgan (2008). The In the following steps we build the workflow components, run the assembled simulation model, and examine the simulated results.

### Step 1: Region

The region of interest for our Hawkweed spread model is the Bogong High Plains area surrounding the ski-resort township of Falls Creek in Victoria, Australia. We derive template raster (GeoTIFF) for the region from the National Vegetation Information System (NVIS) V7.0 (NVIS, 2025) raster layers. We also use the NVIS data in step 2. Users may download [NVIS Raster Geodatabases](https://www.dcceew.gov.au/environment/environment-information-australia/national-vegetation-information-system/data-products) and place them in a suitable directory (e.g. *downloaded_data*) before loading and transforming the NVIS layer. We build our *Region* class object with our 100m resolution template and configure it for two-tier dispersal, whereby dispersal beyond a 1km radius is performed via a courser-grain 500m resolution grid.

```{r example_1, message = FALSE, fig.align = "center"}
# Load NVIS major vegetation groups (MGV)
nvis_rast <- terra::rast(
  paste0("../downloaded_data/NVIS_V7_0_AUST_RASTERS_EXT_ALL/",
         "NVIS_V7_0_AUST_EXT.gdb"))[["NVIS7_0_AUST_EXT_MVG_ALB"]]
# Crop to the region of interest
region_nvis_rast <- terra::crop(nvis_rast,
                                c(1354000, 1372000, -4120000, -4103000))
# Region class object configured with a masked NVIS template
region <- bsspread::Region(+(region_nvis_rast > 0))
region$set_aggr(aggr_factor = 5, inner_radius = 1000)
terra::plot(region$get_template(), colNA = "grey",
            main = "Hawkweed example region")
```

### Step 2: Population model

For our Hawkweed example we will utilise a presence-only population model. We will configure the establishment probability for the population via an approximate mapping of values for vegetation/landscape types provided in Williams, Hahs, & Morgan (2008), to NVIS (2025) major vegetation group (MVG) classes:

| Vegetation type | Est. prob. | Mapped to MVG |
|:-|:-:|:-:|
| Bog | 0.01 | 27 |
| Celmisia sericophylla herbland | 0.01 | 21 |
| Closed heathland | 0.50 | 18 |
| Disturbed areas | 0.99 | 25 |
| Kunzea heathland | 0.50 | 18 |
| Late-lying snowpatch | 0.10 | 26 |
| Open heathland | 0.90 | 18 |
| Poa hiemata tussock grassland | 0.99 | 19 |
| Poa costiniana tussock grassland | 0.99 | 19 |
| Podocarpus heathland | 0.01 | 18 |
| Relic bog | 0.20 | 27 |
| Rocky grassland | 0.50 | 21 |
| Rocky outcrops | 0.01 | 27 |
| Short turf snowpatch | 0.30 | 21 |
| Subalpine grassland | 0.90 | 21 |
| Forests (assumed low) | 0.01 | 3 |
| Woodlands (assumed low) | 0.01 | 5 |

```{r example_2_1, message = FALSE, fig.align = "center"}
# Establishment probability via mapping to NVIS MVG 
establish_pr_rast <- terra::classify(
  region_nvis_rast,
  matrix(c(
    3,  0.01, # Eucalypt Open Forests
    5,  0.01, # Eucalypt Woodlands
    10, 0.01, # Other Forests and Woodlands
    17, mean(c(0.50, 0.50, 0.90, 0.01)), # Other Shrublands (as per Heathlands)
    18, mean(c(0.50, 0.50, 0.90, 0.01)), # Heathlands
    19, 0.99, # Tussock Grasslands
    21, mean(c(0.01, 0.50, 0.30, 0.90)), # Other Grasslands, Herblands, Sedgelands and Rushlands
    24, 0.00, # Inland Aquatic - freshwater, salt lakes, lagoons
    25, 0.99, # Cleared, non-native vegetation, buildings
    26, 0.10, # Unclassified native vegetation
    27, 0.01 # Naturally bare - sand, rock, claypan, mudflat
  ), ncol = 2, byrow = TRUE))
terra::plot(establish_pr_rast, colNA = "grey",
            main = "Hawkweed establishment probability")
```

We can now build our presence-only population model.

```{r example_2_2, message = FALSE, fig.align = "center"}
# Presence-only population model
population_model <- bsspread::PresencePopulation(
  region,
  establish_pr = establish_pr_rast[][,1])
```

### Step 3: Initialisation

Each spread model simulation is initialised with Hawkweed presence in the Falls Creek township area.

```{r example_3, message = FALSE, fig.align = "center"}
# Initialise with Hawkweed presence in Falls Creek township area
initial_rast <- terra::rast("data/falls_creek_town.tif")
initialiser <- bsspread::Initializer(initial_rast,
                                     region = region,
                                     population_model = population_model)
terra::plot(initial_rast, colNA = "grey",
            main = "Initial Hawkweed presence (Falls Creek township)")
```

### Step 4: Dispersal models

### Step 5: Simulator

### Step 6: Results

## References

Andow, D. A., Kareiva, P. M., Levin, S. A., & Okubo, A. (1990). 'Spread of invading organisms'. *Landscape Ecology*, 4(2–3), 177. <doi:10.1007/bf00132860>

Antonov, M., Csárdi, G., Horvát, S., Müller, K., Nepusz, T., Noom, D., Salmon, M., Traag, V., Welles, B. F., & Zanini, F. (2023). 'igraph enables fast and robust network analysis across programming languages'. *arXiv* preprint arXiv:2311.10260. <doi:10.48550/arXiv.2311.10260>

Bossenbroek, J. M., Kraft, C. E., & Nekola, J. C. (2001). 'Prediction of Long-Distance Dispersal Using Gravity Models: Zebra Mussel Invasion of Inland Lakes'. *Ecological Applications*, 11(6), 1778–1788. <doi:10.2307/3061095>

Bradhurst, R., Spring, D., Stanaway, M., Milner, J., & Kompas, T. (2021). 'A generalised and scalable framework for modelling incursions, surveillance and control of plant and environmental pests'. *Environmental Modelling & Software*, 139, N.PAG. <doi:10.1016/j.envsoft.2021.105004>

Carrasco, L. R., Harwood, T. D., Toepfer, S., MacLeod, A., Levay, N., Kiss, J., Baker, R. H. A., Mumford, J. D., & Knight, J. D. (2010-1). 'Dispersal kernels of the invasive alien western corn rootworm and the effectiveness of buffer zones in eradication programmes in Europe'. *Annals of Applied Biology*, 156(1), 63–77. <doi:10.1111/j.1744-7348.2009.00363.x>

Carrasco, L. R., Mumford, J. D., MacLeod, A., Harwood, T., Grabenweger, G., Leach, A. W., Knight, J. D., & Baker, R. H. A. (2010-2). 'Unveiling human-assisted dispersal mechanisms in invasive alien insects: Integration of spatial stochastic simulation and phenology models'. *Ecological Modelling*, 221(17), 2068–2075. <doi:10.1016/j.ecolmodel.2010.05.012>

Crespo-Pérez, V., Rebaudo, F., Silvain, J.-F., & Dangles, O. (2011). 'Modeling invasive species spread in complex landscapes: the case of potato moth in Ecuador'. *Landscape Ecology*, 26(10), 1447. <doi:10.1007/s10980-011-9649-4>

Csárdi, G., Nepusz, T. (2006). 'The igraph software package for complex network research'. *InterJournal*, *Complex Systems*, 1695. <https://igraph.org>

Csárdi, G., Nepusz, T., Traag, V., Horvát, S., Zanini, F., Noom, D., Müller, K., Schoch, D., & Salmon, M. (2026). 'igraph: Network Analysis and Visualization in R'. <doi:10.5281/zenodo.7682609>. R package version 2.3.1, <https://CRAN.R-project.org/package=igraph>

Etherington, T. R. (2016). 'Least-cost modelling and landscape ecology: Concepts, applications, and opportunities'. *Current Landscape Ecology Reports*, 1, 40–53. <doi:10.1007/s40823-016-0006-9>

Fisher, R. A. (1937). 'The wave of advance of advantageous genes'. *Ann. Eugenics* 7, 355–369. <doi:10.1111/j.1469-1809.1937.tb02153.x>

García Adeva, J. J., Botha, J. H., & Reynolds, M. (2012). 'A simulation modelling approach to forecast establishment and spread of Bactrocera fruit flies'. *Ecological Modelling*, 227, 93–108. <doi:10.1016/j.ecolmodel.2011.11.026>

Gippet, J. M., Liebhold, A. M., Fenn-Moltu, G., & Bertelsmeier, C. (2019). 'Human-mediated dispersal in insects'. *Current Opinion in Insect Science*, 35, 96. <doi:10.1016/j.cois.2019.07.005>

Jongejans, E., Skarpaas, O., & Shea, K. (2008). 'Dispersal, demography and spatial population models for conservation and control management'. *Perspectives In Plant Ecology Evolution And Systematics*, 9(3–4), 153–170. <doi:10.1016/j.ppees.2007.09.005>

Lefkovitch Lp. (1965). 'Study of Population Growth in Organisms Grouped by Stages'. *Biometrics*, 21(1), 1–18. <doi:10.2307/2528348>

Leslie, P. H. (1945). 'On the Use of Matrices in Certain Population Mathematics'. *Biometrika*, 33(3), 183–212. <doi:10.2307/2332297>

Muirhead, J. R., Leung, B., Overdijk, C., Kelly, D. W., Nandakumar, K., Marchant, K. R., & MacIsaac, H. J. (2006). 'Modelling local and long-distance dispersal of invasive emerald ash borer Agrilus planipennis (Coleoptera) in North America'. *Diversity & Distributions*, 12(1), 71–79. <doi:10.1111/j.1366-9516.2006.00218.x>

National Vegetation Information System (NVIS) V7.0 <https://www.dcceew.gov.au/> from copyright Commonwealth of Australia 2025

Okubo, A., & Kareiva, P. (2001). 'Some Examples of Animal Diffusion'. In A. Okubo & S. A. Levin (Eds.) *Diffusion and Ecological Problems: Modern Perspectives* (pp. 238-267). Springer New York. <doi:10.1007/9781475749786>

Ricker, W. E. (1958). 'Handbook of computations for biological statistics of fish populations'. *Fisheries Research Board of Canada, Ottawa*, Bulletin No. 119.

Robinet, C., Kehlenbeck, H., Kriticos, D. J., Baker, R. H. A., Battisti, A., Brunel, S., Dupin, M., Eyre, D., Faccoli, M., Ilieva, Z., Kenis, M., Knight, J., Reynaud, P., Yart, A., & van der Werf, W. (2012). 'A Suite of Models to Support the Quantitative Assessment of Spread in Pest Risk Analysis'. *PLoS ONE*, 7(10), 1–18. <doi:10.1371/journal.pone.0043366>

Schneider, D. W., Ellis, C. D., & Cummings, K. S. (1998). 'A Transportation Model Assessment of the Risk to Native Mussel Communities from Zebra Mussel Spread'. *Conservation Biology*, 12(4), 788–800. <doi:10.1111/j.1523-1739.1998.97042.x>

Shaw, M. W. (1995). 'Simulation of Population Expansion and Spatial Pattern when Individual Dispersal Distributions do not Decline Exponentially with Distance'. *Proceedings B: Biological Sciences*, 259(1356), 243–248. <doi:10.1098/rspb.1995.0036>

Skellam, J. G. (1951). 'Random Dispersal in Theoretical Populations'. *Biometrika*, 38(1/2), 196–218. <doi:10.2307/2332328>

Williams, N. S. G., Hahs, A. K., & Morgan, J. W. (2008). 'A Dispersal-Constrained Habitat Suitability Model for Predicting Invasion of Alpine Vegetation'. *Ecological Applications*, 18(2), 347–359. <doi:10.1890/07-0868.1>

Wilson, J. R. U., Dormontt, E. E., Prentis, P. J., Lowe, A. J., & Richardson, D. M. (2009). 'Something in the way you move: dispersal pathways affect invasion success'. *Trends in Ecology & Evolution*, 24(3), 136. <doi:10.1016/j.tree.2008.10.007>

-------------------

ABARES (2024). 'Catchment Scale Land Use of Australia – Update December 2023 version 2', *Australian Bureau of Agricultural and Resource Economics and Sciences*, Canberra, June, CC BY 4.0, <doi:10.25814/2w2p-ph98>

Australian Bureau of Statistics (Reference period: 2015-16 financial year) 'Tourist accommodation - Australia' [data set], Tourist Accommodation, Australia. *ABS*. <https://www.abs.gov.au/statistics/industry/tourism-and-transport/tourist-accommodation-australia/latest-release>

Australian Bureau of Statistics (Reference period: 2016-17 financial year) 'Fertiliser use, Australia, year ended 30 June 2017' [data set], Land Management and Farming in Australia. *ABS*. <https://www.abs.gov.au/statistics/industry/agriculture/land-management-and-farming-australia/latest-release>

Australian Bureau of Statistics (Reference period: 2024-25 financial year) 'Australian population grid 2025 in GeoTIFF format' [data set], Regional population. *ABS*. <https://www.abs.gov.au/statistics/people/population/regional-population/latest-release>

Camac, J. S., Baumgartner, J. B., Robinson, A., & Elith, J. (2020). 'Developing pragmatic maps of establishment likelihood for plant pests'. Tech. Rep. 170607, *Centre of Excellence for Biosecurity Risk Analysis (CEBRA), The University of Melbourne, Melbourne*.
<https://cebra.unimelb.edu.au/__data/assets/pdf_file/0012/3539397/170607_final_report.pdf>

Camac, J. & Baumgartner, J. (2021). '*edmaps* (early detection maps) : An R package for creating Australian maps of establishment likelihood for terrestrial plant pests'. R package <https://github.com/jscamac/edmaps>

Camac, J. S., Baumgartner, J. B., Hester, S., Subasinghe, R., & Collins, S. (2021). 'Using edmaps & Zonation to inform multi-pest early detection surveillance designs'. Tech. Rep. 20121001, *Centre of Excellence for Biosecurity Risk Analysis (CEBRA), The University of Melbourne, Melbourne*. <https://cebra.unimelb.edu.au/__data/assets/pdf_file/0009/3889773/20121001_final_report.pdf>

Camac, J. S. (2024). 'Detect: Designing post-border surveillance schemes. In Hester et al. (Eds.), *Biosecurity: A Systems Perspective*. Taylor & Francis. <doi:10.1201/9781003253204>

Drake, J. M. (2015). 'Range bagging: a new method for ecological niche modelling from presence-only data'. *Journal of the Royal Society Interface*, 12(107), 20150086. <doi:10.1098/rsif.2015.0086>

Fick, S. E., Hijmans, R. J. (2017). 'WorldClim 2: new 1-km spatial resolution climate surfaces for global land areas'. *International Journal of Climatology*, 37, 4302–4315. <doi:10.1002/joc.5086>

GBIF.org (04 May 2026) 'GBIF Occurrence Download'. <doi:10.15468/dl.q6q6fk>


Zizka, A., Silvestro, D., Andermann, T., Azevedo, J., Duarte Ritter, C., Edler, D., Farooq, H., Herdean, A., Ariza, M., Scharn, R., Svanteson, S., Wengstrom, N., Zizka, V., & Antonelli, A. (2019). 'CoordinateCleaner: standardized cleaning of occurrence records from biological collection databases.' *Methods in Ecology and Evolution*, 7. <doi:10.1111/2041-210X.13152>, R package version 3.0.1, <https://github.com/ropensci/CoordinateCleaner>.
